Predicting Enrollment with Stacked Models: Bringing Together Theory and Empirics

### 2021-06-07 Paper/Paper_Draft.Rmd
060621 update

# Methods

@@

I compare the results of these models using the Mean Mean Absolute Error (MASE), Absolute Percent Error (MAPE), and Root Mean Squared Error (RMSE), three widely used measures of model accuracy.

@@

@@

A skeptical reader may justifiably question the value in the above exercise. After all, of course the better specified model predicted more accurately, we *knew* the true parameters and data generating process before starting. This is, however, exactly the point. Often, practitioners approach enrollment prediction from a point of imposed ignorance. Brute force and automated methods of feature selection are treated as substitutes for theoretically informed modeling choices^[This is not to argue that automated and brute force methods of feature selection are inherently bad. Note that the ARIMA model fit in the stacked version above iterates through all combinations of parameters to find the best fit. Such methods are powerful tools but they are best leveraged when informed by theoretically grounded modeling choices.]. In the community college context, where data availability can be limited and the process that generates enrollment data is idiosyncratic in known ways, discarding such knowledge has practical consequences for the accuracy of our enrollment forecasts and consequently, on resource availability. Practitioners ought to take pains to incorporate theoretical knowledge in predictive forecasts. Employing stacked models, as shown above, can be a simple manner of doing so.

>>>>

# Method

@@

I compare the results of these models using the Mean Absolute Error (MASE), Mean Absolute Percent Error (MAPE), and Root Mean Squared Error (RMSE), three widely used measures of predictive accuracy. To ensure results are not the product of random chance, I fit each model to 1000 unique datasets and present the mean value for each measure of model accuracy.

@@

While all tested models predict enrollment with some degree of accuracy, there are large differences between models. Simple linear models without lag terms perform poorest, with MAPE scores near 9%. Simple ARIMA models do not fare much better with MAPE scores hovering near 6%. There are smaller differences between simple linear models with lag terms and stacked models. However, stacked models, both with and without higher order terms, outperform the former, if only by a few tenths of a percent in MAPE.

Though the differences between stacked and linear lagged models are statistically small, their substantive importance should not be dismissed. Such differences can easily mean missing predictions by several hundred students at a community college with a student body in the tens of thousands. Moreover, it should be noted that, while the linear model with a lagged term is more parsimonious than the stacked models, parsimony for its own sake is dubious in circumstances where predictive accuracy is paramount

@@

A skeptical reader may justifiably question the value in the above exercise. After all, of course the better specified model predicted more accurately, we *knew* the true parameters and data generating process before starting. This is, however, exactly the point. Often, practitioners approach enrollment prediction from a point of imposed ignorance. Brute force and automated methods of feature selection are treated as substitutes for theoretically informed modeling choices^[This is not to argue that automated and brute force methods of feature selection are inherently bad. Note that the ARIMA model fit in the stacked version above iterates through all combinations of parameters to find the best fit. Such methods are powerful tools but they are best leveraged when informed by theoretically grounded modeling choices.]. In the community college context, where data availability can be limited and the process that generates enrollment data is idiosyncratic in known ways, discarding such knowledge has practical consequences for the accuracy of our enrollment forecasts and consequently, on resource availability. Practitioners ought to take pains to incorporate theoretical knowledge in predictive forecasts. Employing stacked models, as shown above, can be a simple and effective manner of doing so.
